Polynomial phase estimation by phase unwrapping

### 2013-02-26 paper.tex
spellchecked

\begin{abstract}
Estimating the coefficients of a noisy polynomial phase signal is important in fields including radar, biology and radio communications. One approach attempts to perform polynomial regression on the phase of the signal.  This is complicated by the fact that the phase is \emph{wrapped} modulo $2\pi$ and must be \emph{unwrapped} before regression can be performed.  In this paper we consider an estimator that performs phase unwrapping in a least squares manner.  We call this the \emph{least squares unwrapping} (LSU) estimator.  The LSU estimator can be computed in a reasonable amount of time for data sets of moderate size using existing general purpose algorithms from algebraic number theory.  Under mild conditions on the distribution of the noise we describe the asymptotic properties of this estimator, showing that it is strongly consistent and asymptotically normally distributed.  No `small noise' assumption (common in the literature) is required in our analysis.  As a result the LSU estimator will be accurate, even for very noisy signals, provided that the number of observations is sufficiently large.  A key feature is that the LSU estimator functions accurately over a far wider range of parameters than many popular existing estimators.  Monte-carlo simualtions assert the excellent statistical performance of the LSU estimator when compared to existing state-of-the-art estimators and support our theoretical results.
\end{abstract}

@@

The analysis of most existing polynomial phase estimators make what is known as a `small noise' assumption~\cite{Peleg_linear_FM,Oshea_cpf_2004,Djurovic_haf_cpf_2012,Djuric_phase_unwrap_chirp_1990,Barbarossa_analysis_of_PPS_1997,Barbarossa_PHAF_1998}.  %This amounts to assuming that the estimator $\hat{\mu}_k$ of the $k$th coefficient is such that $N^{k}(\tilde{\mu}_k -  \hat{\mu}_k) \rightarrow 0$ (in some sense) as $N \rightarrow \infty$.  
This assumption is used to justify the use what is called a `perturbation analysis'~\cite{Porat_asympt_HAF_DPT_1996,Peleg_linear_FM,Oshea_cpf_2004,Djurovic_haf_cpf_2012,Djuric_phase_unwrap_chirp_1990,Barbarossa_analysis_of_PPS_1997,Barbarossa_PHAF_1998} to obtain expressions for the asymptotic mean square error of the estimator in question.  The `small noise' assumption is usually justified by Monte-Carlo simulations showing the asymptotic mean square error obtained by the perturbation analysis to be accurate provided that the signal to noise ratio (SNR) is larger than a `threshold' value (see the simulations in Section~\ref{sec:simulations} for examples).  Often the threshold is observed to occur at smaller values of SNR as $N$ increases.  A problem with the `small noise' assumption is that it is not known that the threshold continues to occur at increasingly smaller values of SNR as $N$ increases, or if it instead converges to some finite (but perhaps small) value of SNR.  The consequence of the later being that the estimator will be innaccurate if the SNR is smaller than this finite value, regardless of the number of observations $N$.  Our analysis of the LSU estimator requires no `small noise' assumpion and guarantees that no such finite value of SNR exists for the LSU estimator.  As a result the LSU estimator can be accurate for signals with arbitrarily small SNR provided that the number of observations is sufficiently large.

An analysis of the HAF estimator under the assumption that the noise $\{X_n\}$ is white and Gaussian and without the `small noise' assumption has been made in~\cite{Porat_asympt_HAF_DPT_1996} (see inparticular the proof of Lemma~1 and the remark that follows).  To date, the literature does not contain a corresponding analysis of any estimator based on phase unwrapping.  This has apparently lead to the misconception that phase unwrapping estimators are only accurate when the noise is `small' in some sense.  This misconception would appear to be motivated by early papers such as~\cite{Tretter1985,Kay1989} where it was observed that the noise occuring in the phase of the signal only looks `Gaussian' when the complex noise producing it has small variance.  The results in this paper make it clear that phase unwrapping estimators can be accurate at arbitrarily low SNR.  The information contained in the amplitude is not essential for accurate estimation of polynomial phase signals, and can be discarded with only small loss.  For example, in the case that $\{X_n\}$ is white and Gaussian, our results show that the asymptotic variance of the LSU estimator converges to the Cramer-Rao lower bound (CRB) as the SNR becomes large, and converges to $\tfrac{2}{3}\pi$ times the CRB as the SNR approaches zero.  Related results have been presented previously~\cite{Quinn2009_dasp_phase_only_information_loss}.  It is worth noting that this factor of $\tfrac{2}{3}\pi$ times the CRB at low SNR is better than popular estimators such as the HAF and CPF, for which the ratio between the asymptotic variance and the CRB diverges as the SNR approaches zero~\cite{Porat_asympt_HAF_DPT_1996,Oshea_cpf_2004,Djurovic_haf_cpf_2012}.

A key property of the LSU estimator is that it functions correctly for polynomial phase parameters contained anywhere inside the identifiable region described in Section~\ref{sec:ident_aliasing} and~\cite{Xia_dynamic_range_pps_2001,McKilliam2009IndentifiabliltyAliasingPolyphase}.  This is the maximum possible range of parameters.  By constrast popular estimators such as the HAF and CPF operate correctly only for parameters in a range much smaller than the identifiable region.  This property has been studied previously~\cite{Djurovic_aliasing_2012,McKilliam2009IndentifiabliltyAliasingPolyphase,Zhou_lag_diverity_PPT_1997,Xia_dynamic_range_pps_2001} and can be a critical limitation for some applications, for example, interference supression in the pressence of a jammer~\cite{Djukanovic:2011:PMN:1950994.1951252,Djurovic_aliasing_2012}.  Approaches exist that extend the range of parameters suitable for the HAF~\cite{Zhou_lag_diverity_PPT_1997,Djurovic_aliasing_2012}, but the resulting estimators typically have poor statistical performance.  For this reason, the LSU estimator may be the only presently existing estimator that is viable for certain applications, such as interference supression.  The theoretical results given in Section~\ref{sec:least-squar-unwr} prove that the LSU estimator functions correctly over the entire identifiable region.  To our knowledge this is the first time a result of this type has been obtained for any polynomial phase estimator.

@@

A fundamental problem in lattice theory is the \emph{nearest lattice point problem}~\cite{Agrell2002}. The nearest lattice point problem is: Given $\ybf\in\reals^n$ and some lattice $\Lambda \subset \reals^n$, find a lattice point $\xbf \in \Lambda$ such that the Euclidean distance between $\ybf$ and $\xbf$ is minimised, that is, solve $\arg\min_{\xbf \in \Lambda} \|\ybf - \xbf\|^2.$  The nearest lattice point problem is known to be NP-hard under certain conditions when the lattice itself is considered as an additional input parameter \cite{micciancio_hardness_2001, AjtaiShortestVecProbNPHard1998}.  On the other hand,  fast algorithms that compute a nearest point often exist for specific lattices with known structure~\cite{SPLAG,Conway1982FastQuantDec,McKilliam2009CoxeterLattices,McKilliam2008}.  Even in the general case, where the lattice has no known structure, algorithms exist that can compute a nearest lattice point in reasonable time if the dimension is small~\cite{Agrell2002, Viterbo_sphere_decoder_1999, Pohst_sphere_decoder_1981}. These algorithms typically operate using a tree search to enumerate all lattice points inside a ball of radius sufficiently large to contain a nearest point.  The algorithms vary in the order that branches in the tree are searched.  A popular approach due to Schnorr and Eucher~\cite{schnorr_euchner_sd_1994,Agrell2002} searches branches in a greedy order, and this is the approach we implement here.  We refer to this algorithm as the~\emph{sphere decoder} in this paper, although other variants of the algorithm also go by this name.

With modern computers, the sphere decoder is computationally viable if the dimension of the lattice is less than approximately $60$.  Algorithms that approximate a nearest point have also been studied.  One example is Babai's nearest plane algorithm \cite{Babai1986}.  Another pragmatic approximate approach is to employ the sphere decoder, but prune branches in the search tree to ensure that there are atmost a finite number, say $K$, remaining branches at any time.  This approach has been studied previously~\cite{Baro_list_sphere_dec_2003,Zhan2006_K_best_sphere_decoder,Mondal_Kbest_again_2009} and resembles what is called the sequential $M$-algorithm~\citep{Anderson1984_seq_coding_alg} from coding theory.  The approaches vary in how branches are pruned.  In our implementation we prune branches using the same greedy metric as that used by the Schnorr and Euchner sphere deocder~\cite{schnorr_euchner_sd_1994,Agrell2002} for deciding the order branches are searched.  Following~\cite{Zhan2006_K_best_sphere_decoder} we refer to this as \emph{$K$-best algorithm}.  We will find that this algorithm produces excellent estimates of polynomial phase coefficients and can be computed in a reasonable amount of time for moderate numbers of observations ($N$ less that about 1000).

@@

\section{Proof of strong consistency}\label{sec:strongconstproof}
 Substituting \eqref{eq:noise_circ_poly} into $SS$ we obtain
 \begin{align*}
SS\left( \mubf \right) &= \sum_{n=1}^{N}\fracpart{ \fracpart{ \Phi_n + \sum_{k = 0}^{m}{\tilde{\mu}_k n^k} } - \sum_{k = 0}^{m}{\mu_k n^k} }^{2} \\
&= \sum_{n=1}^{N}\fracpart{  \Phi_n + \sum_{k = 0}^{m}{(\tilde{\mu}_k - \mu_k) n^k} }^{2}.
\end{align*}
Let $\lambdabf = \dealias(\tilde{\mubf} - \mubf) = \tilde{\mubf} - \mubf - \pbf$ where $\pbf$ is a lattice point from $L_{m+1}$. From the definition of $L_{m+1}$ we have $p_0 + p_1 n + \dots + p_{m} n^m$ an integer whenever $n$ is an integer, so
\begin{align*}
\fracpart{\sum_{k=0}^{m}\lambda_k n^k } &= \fracpart{\sum_{k=0}^{m}(\tilde{\mu}_k - \mu_k - p_k) n^k } \\
&= \fracpart{\sum_{k=0}^{m}(\tilde{\mu}_k - \mu_k) n^k }.
\end{align*}
Let
\[
SS\left( \mubf \right) = \sum_{n=1}^{N}\fracpart{  \Phi_n + \sum_{k = 0}^{m}{\lambda_k n^k} }  ^{2} = N S_{N}\left( \lambdabf \right).
 \]
From the definition of the $\dealias(\cdot)$ function $\lambdabf \in B$ so the elements of $\lambdabf$ satisfy
 \begin{equation} \label{eq:identifiability}
 -\frac{0.5}{k!} \leq \lambda_k < \frac{0.5}{k!}.
 \end{equation} 
Now $\widehat{\lambdabf}_N = \dealias(\tilde{\mubf} - \widehat{\mubf})$ is the minimiser of $S_{N}$ in $B$.  %We now analyse the minimiser of $S_N$.  
Let
\[
V_N(\lambdabf) =  \expect S_N(\lambdabf) = \frac{1}{N}\sum_{n=1}^{N} \expect \fracpart{  \Phi_{n}+\sum_{k = 0}^{m}{\lambda_k n^k}}  ^{2}.
\]
Using standard techniques~\cite{Pollard_conv_stat_proc_1984,van2009empirical} one can show that
 \begin{equation}\label{eq:SNVNunifmlln}
\sup_{\lambdabf \in B}\sabs{S_N(\lambdabf) - V_N(\lambdabf)} \rightarrow 0  
 \end{equation}
almost surely as $N\rightarrow\infty$.  This type of resul is called a~\emph{uniform law of large numbers}~\cite{Pollard_conv_stat_proc_1984,van2009empirical}.  A full proof of~\eqref{eq:SNVNunifmlln} in given in~\cite{McKilliam_mean_dir_est_sq_arc_length2010}.  We now concentrate attention on the minimiser of $V_N$. Because $\Phi_n$ has zero intrinsic mean 
\begin{equation}\label{eq:Efracpartmined}
\expect \fracpart{ \Phi_n + z }^{2}
\end{equation}
is minimised uniquely at $z = 0$ for $z \in [-\nicefrac{1}{2}, \nicefrac{1}{2})$.  Since the intrinsic variance of $\Phi_n$ is $\sigma^2$, when $z = 0$,
\begin{equation}\label{eq:Efracpartphi}
\expect\fracpart{\Phi_1+z}^{2} = \expect\fracpart{\Phi_1}^{2} = \sigma^2,
\end{equation}
and so the minimum attained value is $\sigma^2$.

@@

In Figure~\ref{plot:m3small} the CPF estimator reaches a threshold at \unit[1]{dB}, the LSU estimator at \unit[3]{dB} and the HAF estimator at \unit[5]{dB}.  In Figure~\ref{plot:m3small} the true coefficients are $\tilde{\mubf} = \left[ \tfrac{1}{4}, \tfrac{1}{4}, \tfrac{1}{8N}, \tfrac{1}{24N^2}  \right]$.  These are specifically chosen to satisfy the requirements of the HAF and CPF estimators.  The HAF estimator requires that the coefficients satisfy $\tilde{\mu}_0, \tilde{\mu}_1 \in [-0.5, 0.5)$ and
\[
\tilde{\mu}_k \in \left[ -\frac{0.5}{k! \tau^{k-1}}, \frac{0.5}{k! \tau^{k-1}}\right) \qquad k = 2,3,\dots,m,
\]
where $\tau$ is the \emph{lag} and we choose $\tau = \floor{N/m}$ as suggested in~\cite{Peleg_DPT_1995}.  The CPF estimator requires similar, but slightly more complicated conditions on the coeffcients~\cite{Oshea_cpf_2004}.  It is informative to compare the volume of the set of coefficients suitable for the HAF and CPF estimators with the volume of the identifiable region $B$ from~\eqref{eq:rectangular_identifiable_region}. The identifiable region has volume 
\[
\vol(B) = \prod_{k=0}^m {\frac{1}{k!}}.
\]
The volume of coefficients suitable for the HAF estimator is
\[
\prod_{k=2}^m \frac{1}{k! \tau^{k-1}} =  \vol(B) \prod_{k=2}^m \frac{1}{\tau^{k-1}} \leq \vol(B) \prod_{k=2}^m \left( \frac{m}{N} \right)^{k-1}.
\]
The volume of the coefficients suitable for the CPF estimator can be shown to be $\tfrac{3}{2N^3} = 18 \vol(B) N^{-3}$.  It is now worth reconsidering the results in Figure~\ref{plot:m3small}.  Substituting $m=3$ and $N=199$ in the formula above, we see that the HAF and CPF estimators search a space of parameters that is almost $5 \times 10^5$ times smaller than that searched by the LSU estimator.  In this context we might expect the CPF and HAF estimators to have thresholds at lower SNR than LSU estimator.  It is encouraging that the LSU estimator has comparible performance.  To put these results further into context, Figure~\ref{plot:m3big} repeats the simulations in Figure~\ref{plot:m3small}, but where the true coefficients are $\tilde{\mubf} = \left[ \tfrac{1}{4}, \tfrac{1}{4}, \tfrac{1}{8}, \tfrac{1}{24}  \right]$.  The CPF and HAF estimators fail spectacularly.  The performance of the LSU estimator is unchanged, as expected in view of Theorem~\ref{thm:asymp_proof}.

The comparitively small set of coefficients suitable for the HAF and CPF estimators has been studied previously~\cite{Angeby_PPS_aliasing_2000,Djurovic_aliasing_2012,McKilliam2009IndentifiabliltyAliasingPolyphase,Zhou_lag_diverity_PPT_1997,Xia_dynamic_range_pps_2001} and can be a critical limitation for some applications, for example, interference supression in the pressence of a jammer~\cite{Djukanovic:2011:PMN:1950994.1951252,Djurovic_aliasing_2012}.  Zhou and Wang~\cite{Zhou_lag_diverity_PPT_1997} describe a method based on the extended Euclidean algorithm that enables the HAF estimator to operate on the entire identifiable region $B$.  Unfortunately this comes at the price of statistical accuracy.  Figure~\ref{plot:m3bigzw} shows the performance of the Zhou and Wang estimator when the coefficients are $\tilde{\mubf} = \left[ \tfrac{1}{4}, \tfrac{1}{4}, \tfrac{1}{8}, \tfrac{1}{24}  \right]$, the same as in Figure~\ref{plot:m3big}.  The threshold for the estimator is approximately~\unit[29]{dB}.  There does not appear to exist (atleast in the literature) a similar method for increasing the range of coefficients for the CPF estimator.

Figures~\ref{plot:m5} displays the sample MSE of the highest order coefficient $\mu_5$  for a polynomial phase signal of order $m=5$.  In this case the LSU estimator reaches a threshold at \unit[12]{dB}, the HAF estimator at \unit[18]{dB}, and the estimator that combines the CPF and HAF at \unit[10]{dB}.  In this figure the true coefficients are $\tilde{\mubf} = \left[ \tfrac{1}{4}, \tfrac{1}{4}, \tfrac{1}{8N}, \tfrac{1}{24N^2}, \tfrac{1}{96N^3}, \tfrac{1}{480N^4}  \right]$ and these are chosen specifically to satisfy the requirement of the HAF and CPF-HAF estimators.  As in the case where $m=3$, the region of coefficients suitable for the HAF and CPF-HAF estimators is much smaller that the identifiable region $B$.  The effect becomes more servere as $m$ increases.  When $m=5$ and $N=199$ the space of coefficients searched by the HAF and CPF-HAF estimators is approximately $10^{16}$ times smaller than the identifiable region.  This constitutes a dramatic reduction in the range of detectable polynomial phase signals. Figure~\ref{plot:m3big} displays results under the same conditions but with $\tilde{\mubf} = \left[ \tfrac{1}{4}, \tfrac{1}{4}, \tfrac{1}{8}, \tfrac{1}{24}, \tfrac{1}{96}, \tfrac{1}{480}  \right]$.  The conclusion is similar to that from Figure~\ref{plot:m3big}.  The performance of the LSU estimator remains unchanged, but the HAF and CPF-HAF estimators fail completely.

>>>>

\begin{abstract}
Estimating the coefficients of a noisy polynomial phase signal is important in fields including radar, biology and radio communications. One approach attempts to perform polynomial regression on the phase of the signal.  This is complicated by the fact that the phase is \emph{wrapped} modulo $2\pi$ and must be \emph{unwrapped} before regression can be performed.  In this paper we consider an estimator that performs phase unwrapping in a least squares manner.  We call this the \emph{least squares unwrapping} (LSU) estimator.  The LSU estimator can be computed in a reasonable amount of time for data sets of moderate size using existing general purpose algorithms from algebraic number theory.  Under mild conditions on the distribution of the noise we describe the asymptotic properties of this estimator, showing that it is strongly consistent and asymptotically normally distributed.  No `small noise' assumption (common in the literature) is required in our analysis.  As a result the LSU estimator will be accurate, even for very noisy signals, provided that the number of observations is sufficiently large.  A key feature is that the LSU estimator functions accurately over a far wider range of parameters than many popular existing estimators.  Monte-Carlo simulations assert the excellent statistical performance of the LSU estimator when compared to existing state-of-the-art estimators and support our theoretical results.
\end{abstract}

@@

The analysis of most existing polynomial phase estimators make what is known as a `small noise' assumption~\cite{Peleg_linear_FM,Oshea_cpf_2004,Djurovic_haf_cpf_2012,Djuric_phase_unwrap_chirp_1990,Barbarossa_analysis_of_PPS_1997,Barbarossa_PHAF_1998}.  %This amounts to assuming that the estimator $\hat{\mu}_k$ of the $k$th coefficient is such that $N^{k}(\tilde{\mu}_k -  \hat{\mu}_k) \rightarrow 0$ (in some sense) as $N \rightarrow \infty$.  
This assumption is used to justify the use what is called a `perturbation analysis'~\cite{Porat_asympt_HAF_DPT_1996,Peleg_linear_FM,Oshea_cpf_2004,Djurovic_haf_cpf_2012,Djuric_phase_unwrap_chirp_1990,Barbarossa_analysis_of_PPS_1997,Barbarossa_PHAF_1998} to obtain expressions for the asymptotic mean square error of the estimator in question.  The `small noise' assumption is usually justified by Monte-Carlo simulations showing the asymptotic mean square error obtained by the perturbation analysis to be accurate provided that the signal to noise ratio (SNR) is larger than a `threshold' value (see the simulations in Section~\ref{sec:simulations} for examples).  Often the threshold is observed to occur at smaller values of SNR as $N$ increases.  A problem with the `small noise' assumption is that it is not known that the threshold continues to occur at increasingly smaller values of SNR as $N$ increases, or if it instead converges to some finite (but perhaps small) value of SNR.  The consequence of the later being that the estimator will be inaccurate if the SNR is smaller than this finite value, regardless of the number of observations $N$.  Our analysis of the LSU estimator requires no `small noise' assumption and guarantees that no such finite value of SNR exists for the LSU estimator.  As a result the LSU estimator can be accurate for signals with arbitrarily small SNR provided that the number of observations is sufficiently large.

An analysis of the HAF estimator under the assumption that the noise $\{X_n\}$ is white and Gaussian and without the `small noise' assumption has been made in~\cite{Porat_asympt_HAF_DPT_1996} (see in particular the proof of Lemma~1 and the remark that follows).  To date, the literature does not contain a corresponding analysis of any estimator based on phase unwrapping.  This has apparently lead to the misconception that phase unwrapping estimators are only accurate when the noise is `small' in some sense.  This misconception would appear to be motivated by early papers such as~\cite{Tretter1985,Kay1989} where it was observed that the noise occurring in the phase of the signal only looks `Gaussian' when the complex noise producing it has small variance.  The results in this paper make it clear that phase unwrapping estimators can be accurate at arbitrarily low SNR.  The information contained in the amplitude is not essential for accurate estimation of polynomial phase signals, and can be discarded with only small loss.  For example, in the case that $\{X_n\}$ is white and Gaussian, our results show that the asymptotic variance of the LSU estimator converges to the Cramer-Rao lower bound (CRB) as the SNR becomes large, and converges to $\tfrac{2}{3}\pi$ times the CRB as the SNR approaches zero.  Related results have been presented previously~\cite{Quinn2009_dasp_phase_only_information_loss}.  It is worth noting that this factor of $\tfrac{2}{3}\pi$ times the CRB at low SNR is better than popular estimators such as the HAF and CPF, for which the ratio between the asymptotic variance and the CRB diverges as the SNR approaches zero~\cite{Porat_asympt_HAF_DPT_1996,Oshea_cpf_2004,Djurovic_haf_cpf_2012}.

A key property of the LSU estimator is that it functions correctly for polynomial phase parameters contained anywhere inside the identifiable region described in Section~\ref{sec:ident_aliasing} and~\cite{Xia_dynamic_range_pps_2001,McKilliam2009IndentifiabliltyAliasingPolyphase}.  This is the maximum possible range of parameters.  By contrast popular estimators such as the HAF and CPF operate correctly only for parameters in a range much smaller than the identifiable region.  This property has been studied previously~\cite{Djurovic_aliasing_2012,McKilliam2009IndentifiabliltyAliasingPolyphase,Zhou_lag_diverity_PPT_1997,Xia_dynamic_range_pps_2001} and can be a critical limitation for some applications, for example, interference suppression in the presence of a jammer~\cite{Djukanovic:2011:PMN:1950994.1951252,Djurovic_aliasing_2012}.  Approaches exist that extend the range of parameters suitable for the HAF~\cite{Zhou_lag_diverity_PPT_1997,Djurovic_aliasing_2012}, but the resulting estimators typically have poor statistical performance.  For this reason, the LSU estimator may be the only presently existing estimator that is viable for certain applications, such as interference suppression.  The theoretical results given in Section~\ref{sec:least-squar-unwr} prove that the LSU estimator functions correctly over the entire identifiable region.  To our knowledge this is the first time a result of this type has been obtained for any polynomial phase estimator.

@@

A fundamental problem in lattice theory is the \emph{nearest lattice point problem}~\cite{Agrell2002}. The nearest lattice point problem is: Given $\ybf\in\reals^n$ and some lattice $\Lambda \subset \reals^n$, find a lattice point $\xbf \in \Lambda$ such that the Euclidean distance between $\ybf$ and $\xbf$ is minimised, that is, solve $\arg\min_{\xbf \in \Lambda} \|\ybf - \xbf\|^2.$  The nearest lattice point problem is known to be NP-hard under certain conditions when the lattice itself is considered as an additional input parameter \cite{micciancio_hardness_2001, AjtaiShortestVecProbNPHard1998}.  On the other hand,  fast algorithms that compute a nearest point often exist for specific lattices with known structure~\cite{SPLAG,Conway1982FastQuantDec,McKilliam2009CoxeterLattices,McKilliam2008}.  Even in the general case, where the lattice has no known structure, algorithms exist that can compute a nearest lattice point in reasonable time if the dimension is small~\cite{Agrell2002, Viterbo_sphere_decoder_1999, Pohst_sphere_decoder_1981}. These algorithms typically operate using a tree search to enumerate all lattice points inside a ball of radius sufficiently large to contain a nearest point.  The algorithms vary in the order that branches in the tree are searched.  A popular approach due to Schnorr and Euchner~\cite{schnorr_euchner_sd_1994,Agrell2002} searches branches in a greedy order, and this is the approach we implement here.  We refer to this algorithm as the~\emph{sphere decoder} in this paper, although other variants of the algorithm also go by this name.

With modern computers, the sphere decoder is computationally viable if the dimension of the lattice is less than approximately $60$.  Algorithms that approximate a nearest point have also been studied.  One example is Babai's nearest plane algorithm \cite{Babai1986}.  Another pragmatic approximate approach is to employ the sphere decoder, but prune branches in the search tree to ensure that there are at most a finite number, say $K$, remaining branches at any time.  This approach has been studied previously~\cite{Baro_list_sphere_dec_2003,Zhan2006_K_best_sphere_decoder,Mondal_Kbest_again_2009} and resembles what is called the sequential $M$-algorithm~\citep{Anderson1984_seq_coding_alg} from coding theory.  The approaches vary in how branches are pruned.  In our implementation we prune branches using the same greedy metric as that used by the Schnorr and Euchner sphere decoder~\cite{schnorr_euchner_sd_1994,Agrell2002} for deciding the order branches are searched.  Following~\cite{Zhan2006_K_best_sphere_decoder} we refer to this as \emph{$K$-best algorithm}.  We will find that this algorithm produces excellent estimates of polynomial phase coefficients and can be computed in a reasonable amount of time for moderate numbers of observations ($N$ less that about 1000).

@@

\section{Proof of strong consistency}\label{sec:strongconstproof}
 Substituting \eqref{eq:noise_circ_poly} into $SS$ we obtain
 \begin{align*}
SS\left( \mubf \right) &= \sum_{n=1}^{N}\fracpart{ \fracpart{ \Phi_n + \sum_{k = 0}^{m}{\tilde{\mu}_k n^k} } - \sum_{k = 0}^{m}{\mu_k n^k} }^{2} \\
&= \sum_{n=1}^{N}\fracpart{  \Phi_n + \sum_{k = 0}^{m}{(\tilde{\mu}_k - \mu_k) n^k} }^{2}.
\end{align*}
Let $\lambdabf = \dealias(\tilde{\mubf} - \mubf) = \tilde{\mubf} - \mubf - \pbf$ where $\pbf$ is a lattice point from $L_{m+1}$. From the definition of $L_{m+1}$ we have $p_0 + p_1 n + \dots + p_{m} n^m$ an integer whenever $n$ is an integer, so
\begin{align*}
\fracpart{\sum_{k=0}^{m}\lambda_k n^k } &= \fracpart{\sum_{k=0}^{m}(\tilde{\mu}_k - \mu_k - p_k) n^k } \\
&= \fracpart{\sum_{k=0}^{m}(\tilde{\mu}_k - \mu_k) n^k }.
\end{align*}
Let
\[
SS\left( \mubf \right) = \sum_{n=1}^{N}\fracpart{  \Phi_n + \sum_{k = 0}^{m}{\lambda_k n^k} }  ^{2} = N S_{N}\left( \lambdabf \right).
 \]
From the definition of the $\dealias(\cdot)$ function $\lambdabf \in B$ so the elements of $\lambdabf$ satisfy
 \begin{equation} \label{eq:identifiability}
 -\frac{0.5}{k!} \leq \lambda_k < \frac{0.5}{k!}.
 \end{equation} 
Now $\widehat{\lambdabf}_N = \dealias(\tilde{\mubf} - \widehat{\mubf})$ is the minimiser of $S_{N}$ in $B$.  %We now analyse the minimiser of $S_N$.  
Let
\[
V_N(\lambdabf) =  \expect S_N(\lambdabf) = \frac{1}{N}\sum_{n=1}^{N} \expect \fracpart{  \Phi_{n}+\sum_{k = 0}^{m}{\lambda_k n^k}}  ^{2}.
\]
Using standard techniques~\cite{Pollard_conv_stat_proc_1984,van2009empirical} one can show that
 \begin{equation}\label{eq:SNVNunifmlln}
\sup_{\lambdabf \in B}\sabs{S_N(\lambdabf) - V_N(\lambdabf)} \rightarrow 0  
 \end{equation}
almost surely as $N\rightarrow\infty$.  This type of result is called a~\emph{uniform law of large numbers}~\cite{Pollard_conv_stat_proc_1984,van2009empirical}.  A full proof of~\eqref{eq:SNVNunifmlln} in given in~\cite{McKilliam_mean_dir_est_sq_arc_length2010}.  We now concentrate attention on the minimiser of $V_N$. Because $\Phi_n$ has zero intrinsic mean 
\begin{equation}\label{eq:Efracpartmined}
\expect \fracpart{ \Phi_n + z }^{2}
\end{equation}
is minimised uniquely at $z = 0$ for $z \in [-\nicefrac{1}{2}, \nicefrac{1}{2})$.  Since the intrinsic variance of $\Phi_n$ is $\sigma^2$, when $z = 0$,
\begin{equation}\label{eq:Efracpartphi}
\expect\fracpart{\Phi_1+z}^{2} = \expect\fracpart{\Phi_1}^{2} = \sigma^2,
\end{equation}
and so the minimum attained value is $\sigma^2$.

@@

In Figure~\ref{plot:m3small} the CPF estimator reaches a threshold at \unit[1]{dB}, the LSU estimator at \unit[3]{dB} and the HAF estimator at \unit[5]{dB}.  In Figure~\ref{plot:m3small} the true coefficients are $\tilde{\mubf} = \left[ \tfrac{1}{4}, \tfrac{1}{4}, \tfrac{1}{8N}, \tfrac{1}{24N^2}  \right]$.  These are specifically chosen to satisfy the requirements of the HAF and CPF estimators.  The HAF estimator requires that the coefficients satisfy $\tilde{\mu}_0, \tilde{\mu}_1 \in [-0.5, 0.5)$ and
\[
\tilde{\mu}_k \in \left[ -\frac{0.5}{k! \tau^{k-1}}, \frac{0.5}{k! \tau^{k-1}}\right) \qquad k = 2,3,\dots,m,
\]
where $\tau$ is the \emph{lag} and we choose $\tau = \floor{N/m}$ as suggested in~\cite{Peleg_DPT_1995}.  The CPF estimator requires similar, but slightly more complicated conditions on the coefficients~\cite{Oshea_cpf_2004}.  It is informative to compare the volume of the set of coefficients suitable for the HAF and CPF estimators with the volume of the identifiable region $B$ from~\eqref{eq:rectangular_identifiable_region}. The identifiable region has volume 
\[
\vol(B) = \prod_{k=0}^m {\frac{1}{k!}}.
\]
The volume of coefficients suitable for the HAF estimator is
\[
\prod_{k=2}^m \frac{1}{k! \tau^{k-1}} =  \vol(B) \prod_{k=2}^m \frac{1}{\tau^{k-1}} \leq \vol(B) \prod_{k=2}^m \left( \frac{m}{N} \right)^{k-1}.
\]
The volume of the coefficients suitable for the CPF estimator can be shown to be $\tfrac{3}{2N^3} = 18 \vol(B) N^{-3}$.  It is now worth reconsidering the results in Figure~\ref{plot:m3small}.  Substituting $m=3$ and $N=199$ in the formula above, we see that the HAF and CPF estimators search a space of parameters that is almost $5 \times 10^5$ times smaller than that searched by the LSU estimator.  In this context we might expect the CPF and HAF estimators to have thresholds at lower SNR than LSU estimator.  It is encouraging that the LSU estimator has comparable performance.  To put these results further into context, Figure~\ref{plot:m3big} repeats the simulations in Figure~\ref{plot:m3small}, but where the true coefficients are $\tilde{\mubf} = \left[ \tfrac{1}{4}, \tfrac{1}{4}, \tfrac{1}{8}, \tfrac{1}{24}  \right]$.  The CPF and HAF estimators fail spectacularly.  The performance of the LSU estimator is unchanged, as expected in view of Theorem~\ref{thm:asymp_proof}.

The comparatively small set of coefficients suitable for the HAF and CPF estimators has been studied previously~\cite{Angeby_PPS_aliasing_2000,Djurovic_aliasing_2012,McKilliam2009IndentifiabliltyAliasingPolyphase,Zhou_lag_diverity_PPT_1997,Xia_dynamic_range_pps_2001} and can be a critical limitation for some applications, for example, interference suppression in the presence of a jammer~\cite{Djukanovic:2011:PMN:1950994.1951252,Djurovic_aliasing_2012}.  Zhou and Wang~\cite{Zhou_lag_diverity_PPT_1997} describe a method based on the extended Euclidean algorithm that enables the HAF estimator to operate on the entire identifiable region $B$.  Unfortunately this comes at the price of statistical accuracy.  Figure~\ref{plot:m3bigzw} shows the performance of the Zhou and Wang estimator when the coefficients are $\tilde{\mubf} = \left[ \tfrac{1}{4}, \tfrac{1}{4}, \tfrac{1}{8}, \tfrac{1}{24}  \right]$, the same as in Figure~\ref{plot:m3big}.  The threshold for the estimator is approximately~\unit[29]{dB}.  There does not appear to exist (at least in the literature) a similar method for increasing the range of coefficients for the CPF estimator.

Figures~\ref{plot:m5} displays the sample MSE of the highest order coefficient $\mu_5$  for a polynomial phase signal of order $m=5$.  In this case the LSU estimator reaches a threshold at \unit[12]{dB}, the HAF estimator at \unit[18]{dB}, and the estimator that combines the CPF and HAF at \unit[10]{dB}.  In this figure the true coefficients are $\tilde{\mubf} = \left[ \tfrac{1}{4}, \tfrac{1}{4}, \tfrac{1}{8N}, \tfrac{1}{24N^2}, \tfrac{1}{96N^3}, \tfrac{1}{480N^4}  \right]$ and these are chosen specifically to satisfy the requirement of the HAF and CPF-HAF estimators.  As in the case where $m=3$, the region of coefficients suitable for the HAF and CPF-HAF estimators is much smaller that the identifiable region $B$.  The effect becomes more severe as $m$ increases.  When $m=5$ and $N=199$ the space of coefficients searched by the HAF and CPF-HAF estimators is approximately $10^{16}$ times smaller than the identifiable region.  This constitutes a dramatic reduction in the range of detectable polynomial phase signals. Figure~\ref{plot:m3big} displays results under the same conditions but with $\tilde{\mubf} = \left[ \tfrac{1}{4}, \tfrac{1}{4}, \tfrac{1}{8}, \tfrac{1}{24}, \tfrac{1}{96}, \tfrac{1}{480}  \right]$.  The conclusion is similar to that from Figure~\ref{plot:m3big}.  The performance of the LSU estimator remains unchanged, but the HAF and CPF-HAF estimators fail completely.

Figure~\ref{plot:bench} shows the running time of the LSU estimator computed using the $K$-best method versus the HAF estimator for polynomial phase signals of order 3.  The HAF estimator is computed using the fast Fourier transform and has complexity $O(N \log N)$.   The $K$-best method tends to be faster when the SNR is large.  This is due to the algorithm rapidly finding a nearest point, and subsequently terminating branches in its search tree early.  On the other hand, when the SNR is small, all branches in the tree tend to be searched.  Interestingly the LSU estimator computed using the $K$-best algorithm is \emph{faster} than the HAF when the SNR is large and when $N$ is approximately less than 100.  The LSU estimator can be computed in a reasonable amount of time (a few minutes) when $N=1000$.  The algorithm becomes prohibitively expensive for large $N$.  Benchmarks have been performed with polynomial phase signals of order larger than $3$.  The results are omitted due to space constraints.  The running time of the LSU estimator does not appear to change significantly as $m$ increases.  This is expected since a nearest point in a lattice of approximately the same dimension $N-m-1$ must be computed.  The complexity of the HAF estimator grows linearly with $m$.
